---
title: "Abiotic_site_characterization"
output: html_document
date: "2024-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages, set file paths, read in the data

```{r loadings,echo=T,message=FALSE,results='hide'}
#Packages
require(sf)
require(cluster)
require(vegan)
require(raster)
require(factoextra)#for plotting biplots
require(MVN) #significance testing
require(mvnormtest) #significance testing

#file paths
root_folder<-'G:/Shared drives/MacrosystemsBiodiversity/subprojects/MSB_landuse/linear_disturbances'



geofolder<-'G:/Shared drives/MacrosystemsBiodiversity/data/neon_spatial/L0'

sitemeta_folder<-'G:/Shared drives/MacrosystemsBiodiversity/data/landuse/L0'

#data read in
domains<-st_read(file.path(geofolder,'NEON_Domains.shp')) #shape files of domains

sites<-st_read(file.path(geofolder,'NEON_Field_Sites.shp')) #shape files of sites

plots<-st_read(file.path(geofolder,'All_NEON_TOS_Plots_v9/All_NEON_TOS_Plot_Centroids_v9.shp')) #shape files of plots

sitemeta<-read.csv(file.path(sitemeta_folder,'NEON_Field_Site_Metadata_20231026.csv')) #meta data about sites - inc. temp/precip

fragstats<-read.csv(file.path(root_folder,'fragstats_metrics/road_metrics.csv')) #habitat fragmentation data
```

Combine the fragmentation data with the other site data

```{r}
colnames(fragstats)[2]<-'SiteID'

sites3<-merge(fragstats,sites)
```

## First, let's just characterize the fragmentation across habitats.

### This will allow us to determine which indices best characterize the fragmentation of the sites.

Before we do a cluster analysis and PCA, lets check for multivariate normality

```{r}
### prep the data
colnames(sitemeta)[2]<-'SiteID'

clustdat<-merge(sites3,sitemeta,by='SiteID')

### check for normality
clustdat3.us<-clustdat[,c(3:7,9)]

mshapiro.test(t(clustdat3.us))

mvn(clustdat3.us,mvnTest='mardia')
```

P values were less than 0.05 - indicates non normality but that is okay. Scale the data then do a cluster analysis

```{r}
#scale data and add row names
clustdat4<-scale(clustdat3.us)

rownames(clustdat4)<-clustdat$SiteID

#determine the number of clusters
wss<-(nrow(clustdat4)-1)*sum(apply(clustdat4,2,var))

for(i in 2:15){
  wss[i]<-sum(kmeans(clustdat4,center=i)$withinss)
}

plot(1:15,wss,type='b',xlab='number of clusters',ylab='within group sum of squares') #2 clusters probably
```

Do K-means clustering and plot - also MRPP for significance of clusters

```{r}
fit2<-kmeans(clustdat4,centers=2,iter.max=10,nstart=25)

clusplot(clustdat4,fit2$cluster,color=T,shade=T,labels=2)

## Significance testing
#set up data
frag.groups<-fit2$cluster

frag.dist<-vegdist(clustdat4,method='euclidean')
#do the test
set.seed(11)
fragmrpp<-mrpp(frag.dist,frag.groups,permutations=1000)
fragmrpp
```

clusters are significant

## Do a PCA on the data and plot - determine which variables drive the clustering

```{r}
fit2.pc<-princomp(clustdat4,cor=T)
summary(fit2.pc)
fit2.pc$loadings

fviz_pca_biplot(fit2.pc,col.ind=as.character(fit2$cluster),palette=c('purple','orange'))
```

Do significance testing on the loadings

```{r}
#first write a function to do the significane testing - bootstrapping method
sigpca2<-function (x, permutations=1000, ...)
{
  pcnull <- princomp(x, ...)
  res <- pcnull$loadings
  out <- matrix(0, nrow=nrow(res), ncol=ncol(res))
  N <- nrow(x)
  for (i in 1:permutations) {
    pc <- princomp(x[sample(N, replace=TRUE), ], ...)
    pred <- predict(pc, newdata = x)
    r <-  cor(pcnull$scores, pred)
    k <- apply(abs(r), 2, which.max)
    reve <- sign(diag(r[k,]))
    sol <- pc$loadings[ ,k]
    sol <- sweep(sol, 2, reve, "*")
    out <- out + ifelse(res > 0, sol <=  0, sol >= 0)
  }
  out/permutations
}

set.seed(4)

sigpca2(clustdat4, permutations=1000)
```

All loadings are significant but 'edge density' and 'aggregation index' drive PC1 the most - use these in the broader characterization

# Do a cluster analysis on more abiotic factors to see the role of habitat fragmentation for driving patterning of sites

## First, test for normality and prep the data

```{r}

### test for multivariate normality
clustdat2.us<-clustdat[,c(3,5,40,43,44)] #chose only the fragmentation metrics with the highest loadings on PC 1 from the previous analysis

mshapiro.test(t(clustdat2.us)) #pvalue less than 0.5

mvn(clustdat2.us,mvnTest='mardia') #pvalue less than 0.5 - elevation and precip are the belligerents 

#log transform elevation
#log transform precip

clustdat2.us$field_mean_annual_precipitation_mm<-log(clustdat2.us$field_mean_annual_precipitation_mm)
clustdat2.us$field_mean_elevation_m<-log(clustdat2.us$field_mean_elevation_m)

#try tests for normality again

mshapiro.test(t(clustdat2.us)) #p < 0.05

mvn(clustdat2.us,mvnTest='mardia') #p < 0.05, elevation and precip still not normal in univariate space either

#standardize columns to use in cluster analysis

clustdat2<-scale(clustdat[,c(3,5,40,43,44)]) #edge density and aggregation index for fragmentation - strongest PC1 variables from analysis below

rownames(clustdat2)<-clustdat$SiteID
```

Now that we have transformed and set up the data, determine the number of clusters

```{r}
wss<-(nrow(clustdat2)-1)*sum(apply(clustdat2,2,var))

for(i in 2:15){
  wss[i]<-sum(kmeans(clustdat2,center=i)$withinss)
}

plot(1:15,wss,type='b',xlab='number of clusters',ylab='within group sum of squares') #looks like 2 clusters
```

## Do k-means clustering, test significance of clusters, then run a PCA on the data, same as before

```{r}
#cluster analysis
fit1<-kmeans(clustdat2,centers=2,iter.max=10,nstart=25)
clusplot(clustdat2,fit1$cluster,color=T,shade=T,labels=2)

#significance of clusters
envir.groups<-fit1$cluster

envir.dist<-vegdist(clustdat2,method='euclidean')
set.seed(11)
envirmrpp<-mrpp(envir.dist,envir.groups,permutations=1000)
envirmrpp

#PCA + significance testing of loadings
fit1.pc<-princomp(clustdat2)
summary(fit1.pc)
fit1.pc$loadings


fviz_pca_biplot(fit1.pc,col.ind=as.character(fit1$cluster),palette=c('purple','orange'))

set.seed(4)

sigpca2(clustdat4, permutations=1000)
```

Temperature and precipitation are orthogonal to habitat fragmentation - this decoupling has some important implications. For example, biotic patterns not explained by temp/precip at these sites might instead be driven by fragmentation.
